Machine learning topological phases in real space

### 2020-07-22 07_paper/article.tex
First alpha. Will now edit figures of information entropy signatures

%\end{document}

%%%%%% Numerical experiments parameters
\newcommand\nPrec{2}    % precision digits
\newcommand\accPrec{4}  % precision digits for accuracy scores
\newcommand\nHam{6561}  % total number of hamiltonians
\newcommand\epsilonValue{0.01}  % epsilon criterium for test set
\newcommand\nExp{100}   % number of experiments
%%%table commands
%\newcommand{\tableRow}[1]{\\[#1 cm]\hline} %deprecated table row, but might be useful
\newcommand{\tableRow}[1]{\\[#1 cm]}
\newcommand{\tableRowHeader}[1]{\\[#1 cm]\hline}
\newcommand\tableRowHeaderEnd{\tableRowHeader{0.3}}
\newcommand\tableRowEnd{\tableRow{0.15}}
%\newcommand\SSHSys{\thead{SSH systems}}         deprecated, might be useful
%\newcommand\Feat{\thead{Features}}              deprecated, might be useful
%\newcommand\ValEig{\thead{Val. eigenvectors}}   deprecated, might be useful
%\newcommand\TestEig{\thead{Test eigenvectors}}  deprecated, might be useful
%\newcommand\ValHam{\thead{Val. Hamiltonians}}   deprecated, might be useful
%\newcommand\TestHam{\thead{Test Hamiltonians}}  deprecated, might be useful
\newcommand\SSHSys{SSH systems}         
\newcommand\Feat{Features}              
\newcommand\ValEig{Val. eigenvectors}   
\newcommand\TestEig{Test eigenvectors}  
\newcommand\ValHam{Val. Hamiltonians}   
\newcommand\TestHam{Test Hamiltonians}

%%%%%%%%%%%%%%% ssh1 experiments
\newcommand\sshOneTrainNHam{556}
\newcommand\sshOneValNHam{5000}
\newcommand\sshOneTestNHam{1005}
\newcommand\sshOneTrainValWindZeroNHam{2822} %number of winding = 0 in train+val set
\newcommand\sshOneTrainValWindOneNHam{2734}  %number of winding = 1 in train+val set
\newcommand\sshOneTestWindZeroNHam{378}      %number of winding = 0 in test set
\newcommand\sshOneTestWindOneNHam{466}       %number of winding = 1 in test set
%%% real space accuracies
\newcommand\sshOneEigenTrainAcc{0.981438129496403}       % eigenvector training accuracy
\newcommand\sshOneEigenValAcc{0.96392276}         % eigenvector val accuracy
\newcommand\sshOneEigenTestAcc{0.78966836492891}        % eigenvector test accuracy
\newcommand\sshOneHamTrainAcc{1.0}             % Hamiltonian training accuracy
\newcommand\sshOneHamValAcc{1.0}               % Hamiltonian val accuracy
\newcommand\sshOneHamTestAcc{0.9918720379146919}          % Hamiltonian test accuracy

@@

%ssh2
\newcommand\sshTwoTrainNHam{2761}
\newcommand\sshTwoValNHam{2760}
\newcommand\sshTwoTestNHam{1040}

\newcommand\sshTwoTrainValWindMinusOneNHam{988} %number of winding = -1 in train+val set
\newcommand\sshTwoTrainValWindZeroNHam{1795} %number of winding = 0 in train+val set
\newcommand\sshTwoTrainValWindOneNHam{1781}  %number of winding = 1 in train+val set
\newcommand\sshTwoTrainValWindTwoNHam{957}   %number of winding = 2 in train+val set

\newcommand\sshTwoTestWindMinusOneNHam{312} %number of winding = -1 in test set
\newcommand\sshTwoTestWindZeroNHam{95} %number of winding = 0 in test set
\newcommand\sshTwoTestWindOneNHam{109}  %number of winding = 1 in test set
\newcommand\sshTwoTestWindTwoNHam{343}   %number of winding = 2 in test set

@@

%%% x2 features table entries 
\newcommand\xTwo{$X_2$}
\newcommand\xTwoEigTrain{\fpeval{round(0.999655849329953,\accPrec)}}
\newcommand\xTwoEigVal{\fpeval{round(0.9708922101449275,\accPrec)}}
\newcommand\xTwoEigTest{\fpeval{round(0.6634109429569267,\accPrec)}}
\newcommand\xTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xTwoHamVal{\fpeval{round(0.9972499999999997,\accPrec)}}
\newcommand\xTwoHamTest{\fpeval{round(0.879697322467986,\accPrec)}}
%%% x_S2 features table entries  
\newcommand\xSTwo{$X_{\mathcal{S}_2}$}
\newcommand\xSTwoEigTrain{\fpeval{round(0.9996107207533504,\accPrec)}}
\newcommand\xSTwoEigVal{\fpeval{round(0.9589830072463768,\accPrec)}}
\newcommand\xSTwoEigTest{\fpeval{round(0.6168351571594877,\accPrec)}}
\newcommand\xSTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xSTwoHamVal{\fpeval{round(0.9961050724637681,\accPrec)}}
\newcommand\xSTwoHamTest{\fpeval{round(0.8625611175785797,\accPrec)}}
%%% x^c_2 features table entries  
\newcommand\xcTwo{$\hat{X}^c_2$}
\newcommand\xcTwoEigTrain{\fpeval{round(0.9996471930459978,\accPrec)}}
\newcommand\xcTwoEigVal{\fpeval{round(0.9740357971014493,\accPrec)}}
\newcommand\xcTwoEigTest{\fpeval{round(0.6895051222351573,\accPrec)}}
\newcommand\xcTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xcTwoHamVal{\fpeval{round(0.9976050724637681,\accPrec)}}
\newcommand\xcTwoHamTest{\fpeval{round(0.8862281722933641,\accPrec)}}
%%% x^cO_2 features table entries  
\newcommand\xcOTwo{$\hat{X}^c_{\mathcal{O}_2}$}
\newcommand\xcOTwoEigTrain{\fpeval{round(0.99943379210431,\accPrec)}}
\newcommand\xcOTwoEigVal{\fpeval{round(0.8989626449275362,\accPrec)}}
\newcommand\xcOTwoEigTest{\fpeval{round(0.5356883585564609,\accPrec)}}
\newcommand\xcOTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xcOTwoHamVal{\fpeval{round(0.9954927536231883,\accPrec)}}
\newcommand\xcOTwoHamTest{\fpeval{round(0.7896856810244468,\accPrec)}}

%%% x^cSO_2 features table entries  
\newcommand\xcSOTwo{$\hat{X}^c_{\mathcal{S}_2,\mathcal{O}_2}$}
\newcommand\xcSOTwoEigTrain{\fpeval{round(0.9994432452010141,\accPrec)}}
\newcommand\xcSOTwoEigVal{\fpeval{round(0.8998827173913044,\accPrec)}}
\newcommand\xcSOTwoEigTest{\fpeval{round(0.5029968568102445,\accPrec)}}
\newcommand\xcSOTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xcSOTwoHamVal{\fpeval{round(0.9956086956521739,\accPrec)}}
\newcommand\xcSOTwoHamTest{\fpeval{round(0.7670896391152502,\accPrec)}}
%%% x^s_2 features table entries  
\newcommand\xsTwo{$\hat{X}^s_2$}
\newcommand\xsTwoEigTrain{\fpeval{round(0.9996523723288662,\accPrec)}}
\newcommand\xsTwoEigVal{\fpeval{round(0.9735059057971014,\accPrec)}}
\newcommand\xsTwoEigTest{\fpeval{round(0.6878407450523867,\accPrec)}}
\newcommand\xsTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xsTwoHamVal{\fpeval{round(0.9970652173913043,\accPrec)}}
\newcommand\xsTwoHamTest{\fpeval{round(0.8898835855646101,\accPrec)}}
%%% x^SO_2 features table entries  
\newcommand\xsOTwo{$\hat{X}^s_{\mathcal{O}_2}$}
\newcommand\xsOTwoEigTrain{\fpeval{round(0.9994471206084752,\accPrec)}}
\newcommand\xsOTwoEigVal{\fpeval{round(0.9042001086956524,\accPrec)}}
\newcommand\xsOTwoEigTest{\fpeval{round(0.552662980209546,\accPrec)}}
\newcommand\xsOTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xsOTwoHamVal{\fpeval{round(0.9917789855072465,\accPrec)}}
\newcommand\xsOTwoHamTest{\fpeval{round(0.7427124563445866,\accPrec)}}
%%% x^SSO_2 features table entries  
\newcommand\xsSOTwo{$\hat{X}^s_{\mathcal{S}_2,\mathcal{O}_2}$}
\newcommand\xsSOTwoEigTrain{\fpeval{round(0.9992071713147413,\accPrec)}}
\newcommand\xsSOTwoEigVal{\fpeval{round(0.8203919927536232,\accPrec)}}
\newcommand\xsSOTwoEigTest{\fpeval{round(0.41268987194412105,\accPrec)}}
\newcommand\xsSOTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xsSOTwoHamVal{\fpeval{round(0.9941884057971013,\accPrec)}}
\newcommand\xsSOTwoHamTest{\fpeval{round(0.5903026775320139,\accPrec)}}

@@

\textcolor{blue}{The main motivations for developing a data-driven approach based on real space are that wavevector space computations of topological invariants are only possible for systems with translational symmetry, which many physical systems of current interest (e.g. disordered systems in condensed matter) do not have. Moreover, since real space and wavevector space eigenvectors are related by Fourier transforms, the latter are essentially delocalized and therefore so is any information recovered from them. The data-driven approach described in this article addresses these issues. In particular, we employ it to scrutinize the issue of localizability of information in topological condensed matter systems} %\citequote{motif_real_space}

%We develop here an alternative data-driven approach based on real space offers a viable alternative to wavevector space computations that addresses these shortcomings.
%\citequote{motif_real_space}

To investigate topological phases of matter in real space we have designed a novel supervised learning algorithm (here called eigenvector ensembling algorithm) tailored for the task of learning phase transition boundaries from local features. The algorithm is based on eigenvector decomposition and eigenvector ensembling and therefore will require minimal changes to be applicable to a broader class of data-driven physics problems. We demonstrate its effectiveness by combining it with decision trees and random forests to recover the topological phase diagrams of SSH systems from local coordinates of eigenstates in real space.

The advantage of using decision tree-based algorithms to learn topological phases from local eigenvector data is that their use of entropy-based cost functions (such as Shannon information entropy or Gini impurity) furnishes them with an intrinsinc model explanaibility tool that summarizes how important each feature was to learn the desired patterns in the data. This makes it much easier to trace the localization of relevant information along the features of a data set. Here we use the Shannon information entropy of ensembles of real space eigenvectors to recover a signal quantifying the amount of topological information available from each lattice site. This is a highly non-trivial proposition since the topological phase of a system is a global property of the system as a whole emerging from complex interactions between its components, and therefore even defining a local topological signal is a daunting theoretical task. To our knowledge this is the first time that a signal describing the localization of topological information in the bulk of topological condensed matter systems is presented in the literature.

\textcolor{blue}{The topological signals recovered from the analysis of real space eigenvector data are then further theoretically explored as information entropy mass functions along the lattices in real space. By computing their Fourier transform, we retrieve similar information entropy mass functions in wavevector space. These two topological signals, the real space Shannon entropy mass function and its Fourier counterpart in  the reciprocal lattice, are knit together by entropic uncertainty relations. This allows us to establish Heisenberg and Hirschman uncertainty principles for the localizability of Shannon information entropy in topological phase transitions.}  %\citequote{eigenvector_physics}, \citequote{physical_insights}

The Shannon information entropy signals and the corresponding entropic uncertainty relations associated with topological phase transitions presented for the first time in this work provide a clear illustration of how model explainability in machine learning can guide new discoveries in condensed matter and quantum materials physics, since the existence of these signals was established by analyzing data from several thousand SSH systems which, taken individually, could not have provided any concrete hint of their existence.%\citequote{pipeline_not_work}

\textcolor{blue}{As of yet model explainability \cite{gilpin2018explaining}-\cite{roscher2020explainable} is one of the topics at the edge of machine learning research that has been little explored by the physics community working at the interface between the two disciplines. This raises important questions as to whether machine learning can in fact help to advance theoretical investigation in physics, since the majority of physics papers published on the subject are proofs of concept aimed at showing that modern machine learning techniques are capable of recognizing the relevant patterns in data from physical systems whose properties were already known. By proposing new concepts from the analysis of data from physical systems of contemporary interest and knitting together ideas from topological phase transitions and information theory by dint of model explainability, we expect to draw the physics community's attention to this essential machine learning tool.  }%\citequote{pipeline_not_work}
\begin{figure}
\centering
\subfigure[]{\label{ssh1}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh1.png}}\quad
\subfigure[]{\label{ssh2}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh2.png}}
\caption{Phase diagrams in parameter space. a) SSH model with first-neighbor hoppings $t_1$ and $t_2$. The (red) regions with winding number $W$ = 0 are trivial, while the (blue) regions with winding number $W$ = 1 are topologically non-trivial. b) SSH model with first ($t_1$ and $t_2$) and second ($T_1$ and $T_2$) nearest-neighbor hoppings. In this article we set $t_1$ = $t_2$ = 1 and renamed the variables $T_1$ $\rightarrow$ $t_1$, $T_2$ $\rightarrow$ $t_2$ for convenience. The (orange) region with winding number $W$ = 0 is trivial while the others with winding numbers $W$ = -1, $W$ = 1 and $W$ = 2 (red, green and blue respectively) are topologically non-trivial.}
\label{fig:phasediagrams}
\end{figure}

>>>>

%\end{document}
\newcommand{\indicator}[1]{\mathbbm{1}_{\mathcal{#1}}}
%%%%%% Numerical experiments parameters
\newcommand\nPrec{2}    % precision digits
\newcommand\accPrec{4}  % precision digits for accuracy scores
\newcommand\nHam{6561}  % total number of hamiltonians
\newcommand\epsilonValue{0.01}  % epsilon criterium for test set
\newcommand\nExp{100}   % number of experiments
%%%table commands
%\newcommand{\tableRow}[1]{\\[#1 cm]\hline} %deprecated table row, but might be useful
\newcommand{\tableRow}[1]{\\[#1 cm]}
\newcommand{\tableRowHeader}[1]{\\[#1 cm]\hline}
\newcommand\tableRowHeaderEnd{\tableRowHeader{0.3}}
\newcommand\tableRowEnd{\tableRow{0.15}}
%\newcommand\SSHSys{\thead{SSH systems}}         deprecated, might be useful
%\newcommand\Feat{\thead{Features}}              deprecated, might be useful
%\newcommand\ValEig{\thead{Val. eigenvectors}}   deprecated, might be useful
%\newcommand\TestEig{\thead{Test eigenvectors}}  deprecated, might be useful
%\newcommand\ValHam{\thead{Val. Hamiltonians}}   deprecated, might be useful
%\newcommand\TestHam{\thead{Test Hamiltonians}}  deprecated, might be useful
\newcommand\SSHSys{SSH system}         
\newcommand\Feat{Features}              
\newcommand\ValEig{Val. eigenvectors}   
\newcommand\TestEig{Test eigenvectors}  
\newcommand\ValHam{Val. Hamiltonians}   
\newcommand\TestHam{Test Hamiltonians}

%%%%%%%%%%%%%%% ssh1 experiments
%%% data statistics
\newcommand\sshOneTrainNHam{556}
\newcommand\sshOneTrainFracHam{\fpeval{round(\sshOneTrainNHam/\nHam*100,\nPrec)}}
\newcommand\sshOneValNHam{5000}
\newcommand\sshOneValFracHam{\fpeval{round(\sshOneValNHam/\nHam*100,\nPrec)}}
\newcommand\sshOneTrainPlusValNHam{\fpeval{\sshOneTrainNHam+\sshOneValNHam}}
\newcommand\sshOneTestNHam{1005}
\newcommand\sshOneTestFracHam{\fpeval{round(\sshOneTestNHam/\nHam*100,\nPrec)}}
\newcommand\sshOneTrainPlusValWindZeroNHam{2822} %number of winding = 0 in train+val set
\newcommand\sshOneTrainPlusValWindZeroFracHam{\fpeval{round(\sshOneTrainPlusValWindZeroNHam/(\sshOneTrainNHam+\sshOneValNHam)*100,\nPrec)}}
\newcommand\sshOneTrainPlusValWindOneNHam{2734}  %number of winding = 1 in train+val set
\newcommand\sshOneTrainPlusValWindOneFracHam{\fpeval{round(\sshOneTrainPlusValWindOneNHam/(\sshOneTrainNHam+\sshOneValNHam)*100,\nPrec)}}
\newcommand\sshOneTestWindZeroNHam{378}      %number of winding = 0 in test set
\newcommand\sshOneTestWindOneNHam{466}       %number of winding = 1 in test set
\newcommand\sshOneTestWindZeroFracHam{\fpeval{round(\sshOneTestWindZeroNHam/(\sshOneTestWindZeroNHam+\sshOneTestWindOneNHam)*100,\nPrec)}}
\newcommand\sshOneTestWindOneFracHam{\fpeval{round(\sshOneTestWindOneNHam/(\sshOneTestWindZeroNHam+\sshOneTestWindOneNHam)*100,\nPrec)}}
%%% real space accuracies
%\newcommand\sshOneEigenTrainAcc{0.981438129496403}       % eigenvector training accuracy
%\newcommand\sshOneEigenValAcc{0.96392276}         % eigenvector val accuracy
%\newcommand\sshOneEigenTestAcc{0.78966836492891}        % eigenvector test accuracy
%\newcommand\sshOneHamTrainAcc{1.0}             % Hamiltonian training accuracy
%\newcommand\sshOneHamValAcc{1.0}               % Hamiltonian val accuracy
%\newcommand\sshOneHamTestAcc{0.9918720379146919}          % Hamiltonian test accuracy
%\newcommand\sshOneHamTestBaseline{\fpeval{round(\sshOneTestWindOneNHam/(\sshOneTestWindZeroNHam+\sshOneTestWindOneNHam),\accPrec)}}

%%% ssh1 baseline accuracy
\newcommand\sshOneHamTestBaseline{\fpeval{round(\sshOneTestWindOneNHam/(\sshOneTestWindZeroNHam+\sshOneTestWindOneNHam),\accPrec)}}

@@

%ssh2
\newcommand\sshTwoTrainNHam{2761}
\newcommand\sshTwoTrainFracHam{\fpeval{round(\sshTwoTrainNHam/\nHam*100,\nPrec)}}
\newcommand\sshTwoValNHam{2760}
\newcommand\sshTwoValFracHam{\fpeval{round(\sshTwoValNHam/\nHam*100,\nPrec)}}
\newcommand\sshTwoTrainPlusValNHam{\fpeval{\sshTwoTrainNHam+\sshTwoValNHam}}
\newcommand\sshTwoTestNHam{1040}
\newcommand\sshTwoTestFracHam{\fpeval{round(\sshTwoTestNHam/\nHam*100,\nPrec)}}

\newcommand\sshTwoTrainPlusValWindMinusOneNHam{988} %number of winding = -1 in train+val set
\newcommand\sshTwoTrainPlusValWindMinusOneFracHam{\fpeval{round(\sshTwoTrainPlusValWindMinusOneNHam/\sshTwoTrainPlusValNHam*100,\nPrec)}}
\newcommand\sshTwoTrainPlusValWindZeroNHam{1795} %number of winding = 0 in train+val set
\newcommand\sshTwoTrainPlusValWindZeroFracHam{\fpeval{round(\sshTwoTrainPlusValWindZeroNHam/\sshTwoTrainPlusValNHam*100,\nPrec)}}
\newcommand\sshTwoTrainPlusValWindOneNHam{1781}  %number of winding = 1 in train+val set
\newcommand\sshTwoTrainPlusValWindOneFracHam{\fpeval{round(\sshTwoTrainPlusValWindOneNHam/\sshTwoTrainPlusValNHam*100,\nPrec)}}
\newcommand\sshTwoTrainPlusValWindTwoNHam{957}   %number of winding = 2 in train+val set
\newcommand\sshTwoTrainPlusValWindTwoFracHam{\fpeval{round(\sshTwoTrainPlusValWindTwoNHam/\sshTwoTrainPlusValNHam*100,\nPrec)}}

\newcommand\sshTwoTestWindMinusOneNHam{312} %number of winding = -1 in test set
\newcommand\sshTwoTestWindZeroNHam{95} %number of winding = 0 in test set
\newcommand\sshTwoTestWindOneNHam{109}  %number of winding = 1 in test set
\newcommand\sshTwoTestWindTwoNHam{343}   %number of winding = 2 in test set 
\newcommand\sshTwoTestWindMinusOneFracHam{\fpeval{round(\sshTwoTestWindMinusOneNHam/(\sshTwoTestWindMinusOneNHam+\sshTwoTestWindZeroNHam+\sshTwoTestWindOneNHam+\sshTwoTestWindTwoNHam)*100,\nPrec)}}
\newcommand\sshTwoTestWindZeroFracHam{\fpeval{round(\sshTwoTestWindZeroNHam/(\sshTwoTestWindMinusOneNHam+\sshTwoTestWindZeroNHam+\sshTwoTestWindOneNHam+\sshTwoTestWindTwoNHam)*100,\nPrec)}}
\newcommand\sshTwoTestWindOneFracHam{\fpeval{round(\sshTwoTestWindOneNHam/(\sshTwoTestWindMinusOneNHam+\sshTwoTestWindZeroNHam+\sshTwoTestWindOneNHam+\sshTwoTestWindTwoNHam)*100,\nPrec)}}
\newcommand\sshTwoTestWindTwoFracHam{\fpeval{round(\sshTwoTestWindTwoNHam/(\sshTwoTestWindMinusOneNHam+\sshTwoTestWindZeroNHam+\sshTwoTestWindOneNHam+\sshTwoTestWindTwoNHam)*100,\nPrec)}}

@@

%%% ssh2 baseline accuracy
\newcommand\sshTwoHamTestBaseline{\fpeval{round(\sshTwoTestWindTwoNHam/(\sshTwoTestWindMinusOneNHam+\sshTwoTestWindZeroNHam+\sshTwoTestWindOneNHam+\sshTwoTestWindTwoNHam),\accPrec)}}
%%% x2 features table entries 
\newcommand\xTwo{$X_2$}
\newcommand\xTwoEigTrain{\fpeval{round(0.999655849329953,\accPrec)}}
\newcommand\xTwoEigVal{\fpeval{round(0.9708922101449275,\accPrec)}}
\newcommand\xTwoEigTest{\fpeval{round(0.6634109429569267,\accPrec)}}
\newcommand\xTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xTwoHamVal{\fpeval{round(0.9972499999999997,\accPrec)}}
\newcommand\xTwoHamTest{\fpeval{round(0.879697322467986,\accPrec)}}

%%% x_S2 features table entries  
\newcommand\xSTwo{$X_{\mathcal{S}_2}$}
\newcommand\xSTwoEigTrain{\fpeval{round(0.9996107207533504,\accPrec)}}
\newcommand\xSTwoEigVal{\fpeval{round(0.9589830072463768,\accPrec)}}
\newcommand\xSTwoEigTest{\fpeval{round(0.6168351571594877,\accPrec)}}
\newcommand\xSTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xSTwoHamVal{\fpeval{round(0.9961050724637681,\accPrec)}}
\newcommand\xSTwoHamTest{\fpeval{round(0.8625611175785797,\accPrec)}}
%%% x^c_2 features table entries  
\newcommand\xcTwo{$\hat{X}^c_2$}
\newcommand\xcTwoEigTrain{\fpeval{round(0.9996471930459978,\accPrec)}}
\newcommand\xcTwoEigVal{\fpeval{round(0.9740357971014493,\accPrec)}}
\newcommand\xcTwoEigTest{\fpeval{round(0.6895051222351573,\accPrec)}}
\newcommand\xcTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xcTwoHamVal{\fpeval{round(0.9976050724637681,\accPrec)}}
\newcommand\xcTwoHamTest{\fpeval{round(0.8862281722933641,\accPrec)}}
%%% x^cE_2 features table entries  
\newcommand\xcETwo{$\hat{X}^c_{\mathcal{E}_2}$}
\newcommand\xcETwoEigTrain{\fpeval{round(0.99943379210431,\accPrec)}}
\newcommand\xcETwoEigVal{\fpeval{round(0.8989626449275362,\accPrec)}}
\newcommand\xcETwoEigTest{\fpeval{round(0.5356883585564609,\accPrec)}}
\newcommand\xcETwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xcETwoHamVal{\fpeval{round(0.9954927536231883,\accPrec)}}
\newcommand\xcETwoHamTest{\fpeval{round(0.7896856810244468,\accPrec)}}

%%% x^cSE_2 features table entries  
\newcommand\xcSETwo{$\hat{X}^c_{\mathcal{S}_2,\mathcal{E}_2}$}
\newcommand\xcSETwoEigTrain{\fpeval{round(0.9994432452010141,\accPrec)}}
\newcommand\xcSETwoEigVal{\fpeval{round(0.8998827173913044,\accPrec)}}
\newcommand\xcSETwoEigTest{\fpeval{round(0.5029968568102445,\accPrec)}}
\newcommand\xcSETwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xcSETwoHamVal{\fpeval{round(0.9956086956521739,\accPrec)}}
\newcommand\xcSETwoHamTest{\fpeval{round(0.7670896391152502,\accPrec)}}
%%% x^s_2 features table entries  
\newcommand\xsTwo{$\hat{X}^s_2$}
\newcommand\xsTwoEigTrain{\fpeval{round(0.9996523723288662,\accPrec)}}
\newcommand\xsTwoEigVal{\fpeval{round(0.9735059057971014,\accPrec)}}
\newcommand\xsTwoEigTest{\fpeval{round(0.6878407450523867,\accPrec)}}
\newcommand\xsTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xsTwoHamVal{\fpeval{round(0.9970652173913043,\accPrec)}}
\newcommand\xsTwoHamTest{\fpeval{round(0.8898835855646101,\accPrec)}}
%%% x^SO_2 features table entries  
\newcommand\xsOTwo{$\hat{X}^s_{\mathcal{O}_2}$}
\newcommand\xsOTwoEigTrain{\fpeval{round(0.9994471206084752,\accPrec)}}
\newcommand\xsOTwoEigVal{\fpeval{round(0.9042001086956524,\accPrec)}}
\newcommand\xsOTwoEigTest{\fpeval{round(0.552662980209546,\accPrec)}}
\newcommand\xsOTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xsOTwoHamVal{\fpeval{round(0.9917789855072465,\accPrec)}}
\newcommand\xsOTwoHamTest{\fpeval{round(0.7427124563445866,\accPrec)}}
%%% x^SSO_2 features table entries  
\newcommand\xsSOTwo{$\hat{X}^s_{\mathcal{S}_2,\mathcal{O}_2}$}
\newcommand\xsSOTwoEigTrain{\fpeval{round(0.9992071713147413,\accPrec)}}
\newcommand\xsSOTwoEigVal{\fpeval{round(0.8203919927536232,\accPrec)}}
\newcommand\xsSOTwoEigTest{\fpeval{round(0.41268987194412105,\accPrec)}}
\newcommand\xsSOTwoHamTrain{\fpeval{round(1.0,\accPrec)}}
\newcommand\xsSOTwoHamVal{\fpeval{round(0.9941884057971013,\accPrec)}}
\newcommand\xsSOTwoHamTest{\fpeval{round(0.5903026775320139,\accPrec)}}

%%% more feature commands
%\newcommand\xcSE{$\hat{X}^c_{\mathcal{S},\mathcal{E}}$}
%\newcommand\xsSO{$\hat{X}^s_{\mathcal{S},\mathcal{O}}$}

\newcommand\xalpha{$X_\alpha$}
\newcommand\xSalpha{$X_{\mathcal{S}_\alpha}$}

\newcommand\xcalpha{$\hat{X}^c_\alpha$}
\newcommand\xcEalpha{$\hat{X}^c_{\mathcal{E}_\alpha}$}
\newcommand\xcSEalpha{$\hat{X}^c_{\mathcal{S}_\alpha,\mathcal{E}_\alpha}$}

\newcommand\xsalpha{$\hat{X}^s_\alpha$}
\newcommand\xsOalpha{$\hat{X}^s_{\mathcal{O}_\alpha}$}
\newcommand\xsSOalpha{$\hat{X}^s_{\mathcal{S}_\alpha,\mathcal{O}_\alpha}$}

@@

The main motivations for developing a data-driven approach based on real space are that wavevector space computations of topological invariants are only possible for systems with translational symmetry, which many physical systems of current interest (e.g. disordered systems in condensed matter) do not have. Moreover, since real space and wavevector space eigenvectors are related by Fourier transforms, the latter are essentially delocalized and therefore so is any information recovered from them. The data-driven approach described in this article addresses these issues. In particular, we employ it to scrutinize the issue of localizability of information in topological condensed matter systems.

To investigate topological phases of matter in real space we have designed a novel supervised learning algorithm (here called eigenvector ensembling algorithm) tailored for the task of learning phase transition boundaries from local features. The algorithm is based on eigenvector decomposition and eigenvector ensembling and therefore will require minimal changes to be applicable to a broader class of data-driven physics problems. We describe the algorithm in detail in section \ref{the_eigenvector_ensembling_algorithm} and demonstrate its effectiveness by combining it with decision trees and random forests to recover the topological phase diagrams of SSH systems from local coordinates of eigenstates in real space. This is performed in section \ref{numerical_experiments}.

The advantage of using decision tree-based algorithms to learn topological phases from local eigenvector data is that their use of entropy-based cost functions (such as Shannon information entropy or Gini impurity) furnishes them with an intrinsinc model explanaibility tool that summarizes how important each feature was to learn the desired patterns in the data. This makes it much easier to trace the localization of relevant information along the features of a data set. Here we use the Shannon information entropy of ensembles of real space eigenvectors to recover a signal quantifying the amount of topological information available from each lattice site. This is a highly non-trivial proposition since the topological phase of a system is a global property of the system as a whole emerging from complex interactions between its components, and therefore even defining a local topological signal is a daunting theoretical task. To our knowledge this is the first time that a signal describing the localization of topological information in the bulk of topological condensed matter systems is presented in the literature. These topological signals, here called information entropy signatures, are the subject of section \ref{information_entropy_signatures}

In possession of the information entropy signatures, we demonstrate how the symmetries in real and wavevector space eigenvectors can be manipulated with signal processing tools commonly employed in audio and image processing to execute two standard unsupervised learning tasks, namely dimensionality reduction and feature engineering. The topological features resulting from this analysis, here denominated DCT and DST topological features, are discussed in section \ref{topological_feature_engineering}.

The information entropy signatures are finally explored as theoretical constructs in terms of information entropy mass functions along the lattices. By taking their continuum limit and admitting that they are quantum in nature, we show how the information entropy signatures can be naturally understood in terms of emergent information entropy wave functions along the lattices. The theoretical formulation of the results obtained with machine learning in terms of emergent quantum mechanical wave functions allows us to establish Heisenberg and Hirschman uncertainty principles for the localizability of information entropy in topological phase transitions. The emergent information entropy wave functions are the theme of section \ref{emergent_information_entropy_wave_functions}

The  description of information entropy signatures (which were obtained here with machine learning) as a measurable phenomenon emerging from information entropy wave functions provides a clear illustration of how model explainability in machine learning can guide new discoveries in condensed matter and quantum materials physics, since the existence of these signals was established by analyzing data from several thousand SSH systems which, taken individually, could not have provided any concrete hint of their existence.%\citequote{pipeline_not_work}

As of yet model explainability \cite{gilpin2018explaining}-\cite{roscher2020explainable} is one of the topics at the edge of machine learning research that has been little explored by the physics community working at the interface between the two disciplines. This raises important questions as to whether machine learning can in fact help to advance theoretical investigation in physics, since the majority of physics papers published on the subject are proofs of concept aimed at showing that modern machine learning techniques are capable of recognizing the relevant patterns in data from physical systems whose properties were already known. By proposing new concepts from the data analysis of physical systems of contemporary interest and knitting together ideas from topological phase transitions and information theory by dint of model explainability, we expect to draw the physics community's attention to this essential machine learning tool. 
\begin{figure}
\centering
\subfigure[]{\label{ssh1}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh1.png}}\quad
\subfigure[]{\label{ssh2}\includegraphics[width=.45\textwidth]{./phase_diagrams/ssh2.png}}
\caption{Phase diagrams in parameter space. a) SSH model with first-neighbor hoppings $t_1$ and $t_2$. The (red) regions with winding number $W$ = 0 are trivial, while the (blue) regions with winding number $W$ = 1 are topologically non-trivial. b) SSH model with first ($t_1$ and $t_2$) and second ($T_1$ and $T_2$) nearest-neighbor hoppings. In this article we set $t_1$ = $t_2$ = 1 and renamed the variables $T_1$ $\rightarrow$ $t_1$, $T_2$ $\rightarrow$ $t_2$ for convenience. The (orange) region with winding number $W$ = 0 is trivial while the others with winding numbers $W$ = -1, $W$ = 1 and $W$ = 2 (red, green and blue respectively) are topologically non-trivial.}
\label{fig:phasediagrams}
\end{figure}
